---
title: "Thursday - Jumpstart 2025: Data Visualization"
author: "Hannah Overbye-Thompson"
date: "2025-08-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(here)
library(tidyverse)
library(ggplot2)
```

# The Power of Data Visualization in Research

Welcome to data visualization! Today you'll learn to create compelling, publication-ready visualizations that communicate your research findings effectively.

## Why Data Visualization Matters

Data visualization is crucial for communication researchers because it allows us to:

-   **Communicate findings** to academic and public audiences
-   **Identify patterns** and trends in complex datasets\
-   **Spot outliers** and unexpected relationships
-   **Compare groups** and experimental conditions
-   **Tell stories** with data that support theoretical arguments

## Today's Learning Goals:

-   Introduce the grammar of graphics with ggplot2
-   Create publication-ready visualizations for communication research
-   Learn best practices for academic data visualization
-   Practice!

------------------------------------------------------------------------

# Loading Our Data

```{r}
# Load our communication datasets
media_trust <- read_csv(here("data", "media_trust_survey.csv"))
social_media <- read_csv(here("data", "social_media_engagement.csv"))
news_consumption <- read_csv(here("data", "news_consumption_patterns.csv"))

# Quick peek at our data
glimpse(media_trust)
```

------------------------------------------------------------------------

# The Grammar of Graphics: ggplot2 Fundamentals

ggplot2 works on the principle of "grammar of graphics" - building visualizations layer by layer. Think of it like constructing a sentence with specific grammatical rules (see more [here](https://ggplot2-book.org/mastery.html) 🔗)

## Basic Structure

Every ggplot has three essential components: 1. **Data** - your dataset 2. **Aesthetics (aes)** - what variables to plot and how 3. **Geometries (geom)** - the type of plot

```{r}
# Start with an empty plot canvas
ggplot(data = media_trust)
```

```{r}
# Add aesthetic mappings (but still no visual elements)
ggplot(data = media_trust, aes(x = age, y = trust_score))
```

```{r}
# Add a geometry to actually see the data
ggplot(data = media_trust, aes(x = age, y = trust_score)) +
  geom_point()
```

------------------------------------------------------------------------

# Essential Plot Types for Communication Research

## 1. Scatter Plots: Exploring Relationships

**When to use:** Examining relationships between two continuous variables

**Research Example:** "How does age relate to trust in news sources?"

```{r}
# Basic scatter plot
ggplot(media_trust, aes(x = age, y = trust_score)) +
  geom_point()
```

### Adding Layers and Aesthetics

```{r}
# Add color by news source and a trend line
ggplot(media_trust, aes(x = age, y = trust_score)) +
  geom_point(aes(color = news_source), alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    title = "Age and Trust in News Sources",
    subtitle = "Higher trust scores indicate greater trust in news media",
    x = "Age (years)",
    y = "Trust Score (1-7 scale)",
    color = "News Source"
  ) +
  theme_minimal()
```

### Publication-Ready Styling

```{r}
# Professional academic plot
ggplot(media_trust, aes(x = age, y = trust_score)) +
  geom_point(aes(color = news_source), alpha = 0.6, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue", linewidth = 1.2) +
  scale_color_brewer(type = "qual", palette = "Set2") +
  labs(
    title = "Relationship Between Age and News Trust Across Media Sources",
    x = "Participant Age",
    y = "Trust Score",
    color = "Primary News Source",
    caption = "Data: Media Trust Survey 2025 (N = 250)"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
```

## 2. Bar Plots: Comparing Categories

**When to use:** Comparing frequencies or means across categorical variables

### Simple Frequency Bar Plot

```{r}
# Count of participants by news source
ggplot(media_trust, aes(x = news_source)) +
  geom_bar(fill = "steelblue", alpha = 0.8) +
  labs(
    title = "Distribution of Primary News Sources",
    x = "News Source",
    y = "Number of Participants"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Column Plot with Calculated Means

```{r}
# Calculate mean trust by news source
trust_by_source <- media_trust %>%
  group_by(news_source) %>%
  summarize(
    mean_trust = mean(trust_score, na.rm = TRUE),
    se_trust = sd(trust_score, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

# Create column plot with error bars
ggplot(trust_by_source, aes(x = reorder(news_source, mean_trust), y = mean_trust)) +
  geom_col(fill = "darkblue", alpha = 0.7) +
  geom_errorbar(aes(ymin = mean_trust - se_trust, ymax = mean_trust + se_trust), 
                width = 0.2, alpha = 0.8) +
  labs(
    title = "Mean Trust Scores by News Source",
    x = "News Source",
    y = "Mean Trust Score",
    caption = "Error bars represent standard error"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 3. Histograms: Understanding Distributions

**When to use:** Examining the distribution of a single continuous variable

```{r}
# Distribution of trust scores
ggplot(media_trust, aes(x = trust_score)) +
  geom_histogram(bins = 20, fill = "lightblue", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(trust_score, na.rm = TRUE)), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Distribution of News Trust Scores",
    subtitle = "Red line indicates sample mean",
    x = "Trust Score",
    y = "Frequency"
  ) +
  theme_minimal()
```

### Comparing Distributions by Group

```{r}
# Overlapping histograms by gender
ggplot(media_trust %>% 
         filter(gender %in% c("Male", "Female")), 
       aes(x = trust_score, fill = gender)) +
  geom_histogram(bins = 15, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("Male" = "lightblue", "Female" = "pink")) +
  labs(
    title = "Distribution of Trust Scores by Gender",
    x = "Trust Score",
    y = "Frequency",
    fill = "Gender"
  ) +
  theme_minimal()
```

## 4. Box Plots: Comparing Group Distributions

**When to use:** Comparing distributions across multiple groups

```{r}
# Trust scores by education level
ggplot(media_trust, aes(x = education, y = trust_score)) +
  geom_boxplot(aes(fill = education), alpha = 0.7) +
  scale_fill_brewer(type = "qual", palette = "Pastel1") +
  labs(
    title = "Trust Score Distributions by Education Level",
    x = "Education Level",
    y = "Trust Score"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"  # Remove redundant legend
  )
```

## 5. Faceted Plots: Multiple Subplots

**When to use:** Comparing patterns across different groups or conditions

```{r}
# Scatter plot faceted by news source
ggplot(media_trust, aes(x = age, y = trust_score)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~news_source, ncol = 3) +
  labs(
    title = "Age-Trust Relationship Across News Sources",
    x = "Age",
    y = "Trust Score"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

## 6. Advanced: Violin Plots and Other Distributions

```{r}
# Violin plot for rich distribution visualization
ggplot(media_trust, aes(x = education, y = trust_score, fill = education)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.1, alpha = 0.8) +
  scale_fill_viridis_d() +
  labs(
    title = "Trust Score Distributions by Education (Violin Plot)",
    subtitle = "Violin shape shows distribution density, box shows quartiles",
    x = "Education Level",
    y = "Trust Score"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

------------------------------------------------------------------------

# Creating Academic-Quality Visualizations

## Best Practices for Publication

### 1. Clear Titles and Labels

```{r}
# Example of proper academic labeling
ggplot(media_trust, aes(x = political_interest, y = trust_score)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm") +
  labs(
    title = "Political Interest and Trust in News Media",
    subtitle = "Linear relationship between political engagement and media trust",
    x = "Political Interest (1 = Not at all interested, 7 = Very interested)",
    y = "News Trust Score (1 = Very low trust, 7 = Very high trust)",
    caption = "Source: Media Trust Survey 2025; N = 250; r = .34, p < .001"
  ) +
  theme_classic()
```

### 2. Appropriate Color Schemes

```{r}
# Using colorblind-friendly palette
ggplot(media_trust, aes(x = age, y = trust_score, color = gender)) +
  geom_point(alpha = 0.7) +
  scale_color_viridis_d(option = "plasma", end = 0.8) +  # Colorblind friendly
  labs(
    title = "Age, Gender, and News Trust",
    x = "Age",
    y = "Trust Score",
    color = "Gender"
  ) +
  theme_minimal()
```

### 3. Professional Themes

```{r}
# Custom theme for consistency
academic_theme <- theme_classic() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.5),
    panel.grid.minor = element_blank()
  )

# Apply to a plot
ggplot(media_trust, aes(x = daily_news_minutes, y = trust_score)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Daily News Consumption and Trust Levels",
    x = "Daily News Consumption (minutes)",
    y = "Trust Score"
  ) +
  academic_theme
```

------------------------------------------------------------------------

# Practice Exercises with mtcars Dataset

Now let's practice with the classic `mtcars` dataset to reinforce these concepts! Here is a [cheat sheet](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf) to help.

If you need help, please put up your red sticky note.

```{r}
# Load the mtcars dataset
data("mtcars")

# Convert cylinder count to a factor for better plotting
mtcars$cyl <- as.factor(mtcars$cyl)

# Quick exploration
glimpse(mtcars)
```

## Exercise 1: Basic Scatter Plot

Create a scatter plot examining the relationship between weight (`wt`) and fuel efficiency (`mpg`):

### 1a. Basic scatter plot

```{r}
# Your turn: Create a basic scatter plot with wt on x-axis and mpg on y-axis
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()
```

### 1b. Add color by cylinder count

```{r}
# Add color coding by number of cylinders
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(color = cyl), size = 2)
```

### 1c. Add trend line

```{r}
# Add a regression line to show the overall relationship
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(color = cyl), size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "black")
```

### 1d. Professional labels and styling

```{r}
# Your code here: Add proper titles, labels, and professional styling

```

### 1e. Interpretation

**Write a brief interpretation of your plot here:**

*[Your interpretation of the weight-mpg relationship]*

## Exercise 2: Bar Plot of Cylinder Distribution

```{r}
# Create a bar plot showing the frequency of cars by cylinder count
ggplot(mtcars, aes(x = cyl)) +
  geom_bar(fill = "steelblue", alpha = 0.8) +
  labs(
    title = "Distribution of Cars by Cylinder Count",
    x = "Number of Cylinders",
    y = "Number of Cars"
  ) +
  theme_minimal()
```

## Exercise 3: Mean Comparison Plot

### 3a. Calculate average quarter-mile time by cylinder count

```{r}
# Calculate mean quarter-mile time (qsec) by cylinder count
average_qsec <- mtcars %>%
  group_by(cyl) %>%
  summarize(
    mean_qsec = mean(qsec, na.rm = TRUE),
    se_qsec = sd(qsec, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

print(average_qsec)
```

### 3b. Create column plot with error bars

```{r}
# Plot average quarter-mile time by cylinder count
ggplot(average_qsec, aes(x = cyl, y = mean_qsec)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  geom_errorbar(aes(ymin = mean_qsec - se_qsec, ymax = mean_qsec + se_qsec),
                width = 0.2) +
  labs(
    title = "Average Quarter-Mile Time by Cylinder Count",
    x = "Number of Cylinders", 
    y = "Quarter-Mile Time (seconds)",
    caption = "Error bars represent standard error"
  ) +
  theme_classic()
```

### 3c. Interpretation

**Write your interpretation here:**

*[Your interpretation of the cylinder-performance relationship]*

## Exercise 4: Advanced Challenge

Create a comprehensive visualization that shows the relationship between horsepower (`hp`) and quarter-mile time (`qsec`), colored by transmission type (`am` where 0 = automatic, 1 = manual):

```{r}
# First, create a better transmission variable
mtcars <- mtcars %>%
  mutate(transmission = ifelse(am == 0, "Automatic", "Manual"))

# Your code here: Create an advanced plot with:
# - Scatter plot of hp vs qsec
# - Color by transmission type  
# - Separate trend lines for each transmission type
# - Professional styling and labels
# - Proper title and caption


```

## Exercise 5: Faceted Analysis

Create a faceted plot showing the relationship between weight and fuel efficiency, separated by cylinder count:

```{r}
# Your code here: Create faceted scatter plots


```

------------------------------------------------------------------------

# Saving and Exporting Plots

For academic publications, you'll need to save high-quality versions of your plots:

```{r eval=FALSE}
# Save plot as high-resolution PNG
ggsave("trust_by_age.png", width = 8, height = 6, dpi = 300)

# Save as PDF for publications  
ggsave("trust_by_age.pdf", width = 8, height = 6)

# Save as SVG for web use
ggsave("trust_by_age.svg", width = 8, height = 6)
```

------------------------------------------------------------------------

# Key Principles for Research Visualizations

## 1. **Clarity First**

-   Clear visualizations help audiences understand patterns in data
-   Always prioritize understanding over aesthetics
-   Use descriptive titles and axis labels

## 2. **Appropriate Chart Types**

-   **Scatter plots**: Relationships between continuous variables
-   **Bar/Column plots**: Comparing categories or group means
-   **Box plots**: Comparing distributions across groups
-   **Histograms**: Understanding single variable distributions

## 3. **Honest Representation**

-   Don't manipulate scales to exaggerate effects
-   Include confidence intervals and error bars when appropriate
-   Clearly indicate sample sizes and data sources
-   Make sure your axes start at 0 (or another reasonable point), and are the same across graphs

## 4. **Accessibility**

-   Use colorblind-friendly palettes
-   Provide alternative text descriptions
-   Ensure sufficient contrast

## 5. **Publication Standards**

-   Follow your target journal's figure guidelines
-   Include proper statistical information in captions
-   Use consistent styling across all figures

## 6. **An Example** 

-   <https://academic.oup.com/jcmc/article/29/1/zmad039/7394122>

------------------------------------------------------------------------

# What You've Accomplished Today

Excellent work! You've now mastered:

✅ **ggplot2 fundamentals** - grammar of graphics approach

✅ **Essential plot types** for research

✅ **Professional styling** for academic publications

✅ **Best practices** for honest, accessible visualization

## Your Visualization Toolkit:

-   Scatter plots for relationships
-   Bar plots for comparisons
-   Histograms for distributions
-   Box plots for group comparisons
-   Faceted plots for complex analyses
-   Professional themes and styling

## Resources for Continued Learning:

-   [ggplot2 Reference](https://ggplot2.tidyverse.org/reference/)
-   [R Graphics Cookbook](https://r-graphics.org/)
-   [Data Visualization Best Practices](https://www.tableau.com/learn/articles/data-visualization)
-   [Colorbrewer Palettes](https://colorbrewer2.org/)
-   [Academic Figure Guidelines](https://www.nature.com/nature/for-authors/formatting-guide)

Great job building your data visualization skills! 📊✨
