---
title: "Wednesday - Jumpstart 2025: Statistics in R"
author: "Hannah Overbye-Thompson"
date: "2025-08-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(broom)  # For tidy statistical output

# Turn off scientific notation for easier reading
options(scipen = 999)
```

# Welcome Back to Jumpstart 2025!

Today we'll learn how to conduct statistical analyses in R using your communication research data. We'll cover the most common statistical tests used in communication.

## Today's Learning Goals:
- Understand when to use different statistical tests
- Learn to run and interpret statistical analyses in R
- Practice with real communication research scenarios
- Develop skills in hypothesis testing and interpretation

---

# Loading Our Communication Data

Let's start by loading the datasets we'll use for today's analyses:

```{r}
# Load our communication research datasets
media_trust <- read_csv(here("data", "media_trust_survey.csv"))
social_media <- read_csv(here("data", "social_media_engagement.csv"))
news_consumption <- read_csv(here("data", "news_consumption_patterns.csv"))
```

## Quick Data Exploration

Before running any statistics, always explore your data first:

```{r}
# Explore the media trust dataset
glimpse(media_trust)
summary(media_trust$trust_score)
summary(media_trust$age)

# Check for missing values
sum(is.na(media_trust$trust_score))

# Look at the distribution of categorical variables
table(media_trust$news_source)
table(media_trust$gender)
```

---

# Statistical Tests in Communication Research

## 1. One-Sample t-Test

**When to use:** Test if a sample mean differs from a known population value or theoretical expectation.

**Research Question:** "Do participants' trust scores differ significantly from the scale midpoint (4.0)?"

```{r}
# One-sample t-test: Is trust score significantly different from midpoint?
one_sample_result <- t.test(media_trust$trust_score, mu = 4.0)
one_sample_result
```

## 2. Independent Samples t-Test

**When to use:** Compare means between two independent groups.

**Research Question:** "Do males and females differ in their trust scores?"

```{r}
# Filter for just male and female participants
gender_comparison <- media_trust %>%
  filter(gender %in% c("Male", "Female"))

# Independent samples t-test
independent_t_result <- t.test(trust_score ~ gender, data = gender_comparison)
independent_t_result

# Descriptive statistics by group
gender_comparison %>%
  group_by(gender) %>%
  summarize(
    mean_trust = mean(trust_score, na.rm = TRUE),
    sd_trust = sd(trust_score, na.rm = TRUE),
    n = n()
  )
```

## 3. Paired Samples t-Test

**When to use:** Compare two related measurements from the same participants.

**Research Question:** "Do trust scores and credibility ratings differ significantly?"

```{r}
# Paired samples t-test: trust vs credibility ratings
paired_t_result <- t.test(media_trust$trust_score, 
                         media_trust$credibility_rating, 
                         paired = TRUE) 
paired_t_result

# Descriptive statistics
cat("Mean trust score:", round(mean(media_trust$trust_score, na.rm = TRUE), 2), "\n")
cat("Mean credibility rating:", round(mean(media_trust$credibility_rating, na.rm = TRUE), 2))
```

## 4. Linear Regression

**When to use:** Predict one continuous variable from another continuous variable.

**Research Question:** "Does age predict trust in news sources?"

```{r}
# Simple linear regression
trust_age_model <- lm(trust_score ~ age, data = media_trust)
summary(trust_age_model)
```

## 5. Multiple Regression

**When to use:** Predict one variable from multiple predictors.

**Research Question:** "Do age, political interest, and daily news consumption predict trust scores?"

```{r}
# Multiple regression
multiple_model <- lm(trust_score ~ age + political_interest + daily_news_minutes, 
                    data = media_trust)

summary(multiple_model)
```

## 6. Correlation Analysis

**When to use:** Examine the strength and direction of relationships between variables.

**Research Question:** "How strongly are trust scores and credibility ratings related?"

```{r}
# Pearson correlation
correlation_result <- cor.test(media_trust$trust_score, 
                              media_trust$credibility_rating)

print(correlation_result)
```

## 7. One-Way ANOVA

**When to use:** Compare means across three or more groups.

**Research Question:** "Do trust scores differ across education levels?"

```{r}
# One-way ANOVA
anova_model <- aov(trust_score ~ education, data = media_trust)
summary(anova_model)

```

## 8. Two-Way ANOVA

**When to use:** Examine effects of two categorical variables and their interaction.

**Research Question:** "Do trust scores vary by gender and education level?"

```{r}
# Filter for main gender categories for cleaner analysis
anova_data <- media_trust %>%
  filter(gender %in% c("Male", "Female")) %>%
  filter(education %in% c("Bachelor's", "Master's", "PhD"))

# Two-way ANOVA
twoway_model <- aov(trust_score ~ gender * education, data = anova_data)
summary(twoway_model)

# Means by group
anova_data %>%
  group_by(gender, education) %>%
  summarize(
    mean_trust = mean(trust_score, na.rm = TRUE),
    n = n(),
    .groups = 'drop'
  )
```

## 9. Chi-Square Goodness of Fit

**When to use:** Test if observed frequencies match expected frequencies.

**Research Question:** "Are news sources equally preferred among participants?"

```{r}
# Chi-square goodness of fit test
news_source_table <- table(media_trust$news_source)
news_source_table

chisq_goodness <- chisq.test(news_source_table)
chisq_goodness


# Expected vs observed frequencies
data.frame(
  News_Source = names(news_source_table),
  Observed = as.numeric(news_source_table),
  Expected = round(chisq_goodness$expected, 1)
)
```

## 10. Chi-Square Test of Independence

**When to use:** Test if two categorical variables are independent.

**Research Question:** "Is there an association between gender and preferred news source?"

```{r}
# Create contingency table
contingency_table <- table(media_trust$gender, media_trust$news_source)
contingency_table

# Chi-square test of independence
chisq_independence <- chisq.test(contingency_table)
chisq_independence

```

## 11. Logistic Regression

**When to use:** Predict a binary outcome from one or more predictors.

**Research Question:** "What factors predict high trust (above median) in news sources?"

```{r}
# Create binary outcome variable
media_trust <- media_trust %>%
  mutate(high_trust = ifelse(trust_score > median(trust_score, na.rm = TRUE), 1, 0))

# Logistic regression
logistic_model <- glm(high_trust ~ age + political_interest + daily_news_minutes, 
                     data = media_trust, 
                     family = binomial)

summary(logistic_model)
```

---

# Practice Exercises: Your Turn!

Now it's time to practice with our social media engagement dataset!

## Data Exploration (Always Start Here!)

```{r}
# Explore the social media dataset
glimpse(social_media)

# Your task: Calculate descriptive statistics for 3 variables
# 1. One continuous variable (e.g., engagement_score)
# 2. One categorical variable (e.g., platform)
# 3. One of your choice

# Your code here:




```

## Exercise 1: Independent Samples t-Test

**Research Question:** "Do users who post frequently (>4 posts/week) have different engagement scores than those who post less frequently?"

```{r}
# Step 1: Create a binary grouping variable
# Step 2: Run an independent samples t-test
# Step 3: Calculate descriptive statistics by group

# Your code here:




```

## Exercise 2: Correlation Analysis

**Research Question:** "What is the relationship between daily hours spent on social media and engagement scores?"

```{r}
# Step 1: Run a correlation test

# Your code here:




```

## Exercise 3: One-Way ANOVA

**Research Question:** "Do engagement scores differ significantly across social media platforms?"

```{r}
# Step 1: Run a one-way ANOVA
# Step 2: Calculate descriptive statistics by platform

# Your code here:




```

## Exercise 4: Linear Regression

**Research Question:** "Does follower count predict engagement scores?"

```{r}
# Step 1: Run a simple linear regression
# Step 2: Examine model fit statistics

# Your code here:




```

## Exercise 5: Chi-Square Test

**Research Question:** "Is there an association between platform type and news sharing frequency (high vs. low)?"

```{r}
# Step 1: Create a binary news sharing variable (above/below median)
# Step 2: Create a contingency table
# Step 3: Run a chi-square test of independence

# Your code here:




```

## Exercise 6: Multiple Regression

**Research Question:** "What combination of factors (age, daily hours, follower count) best predicts engagement scores?"

```{r}
# Step 1: Run a multiple regression
# Step 2: Examine which predictors are significant

# Your code here:




```

## Exercise 7: Advanced Challenge

**Research Question:** Design your own analysis! Choose variables from any of our datasets and:

1. State a clear research question
2. Choose the appropriate statistical test
3. Run the analysis
4. Interpret the results

```{r}
# Your research question: 


# Justification for your chosen test:


# Your analysis:




# Your interpretation:




```

---

# Interpreting Statistical Results

## Key Guidelines for Communication Research:

### P-values:
- **p < 0.001**: Very strong evidence against null hypothesis
- **p < 0.01**: Strong evidence against null hypothesis  
- **p < 0.05**: Moderate evidence against null hypothesis
- **p ≥ 0.05**: Insufficient evidence to reject null hypothesis

### Effect Sizes:
- **Cohen's d** (t-tests): 0.2 (small), 0.5 (medium), 0.8 (large)
- **R-squared** (regression): % of variance explained
- **Eta-squared** (ANOVA): proportion of variance explained
- **Cramér's V** (chi-square): 0.1 (small), 0.3 (medium), 0.5 (large)

### Reporting Results:
Always include:
1. Descriptive statistics (means, standard deviations)
2. Test statistic and degrees of freedom
3. P-value
4. Effect size
5. Confidence intervals when appropriate

---

# Best Practices for Statistical Analysis

## Before Running Tests:
1. **Explore your data** - check distributions, outliers, missing values
2. **Check assumptions** - normality, homogeneity of variance, independence
3. **Visualize relationships** - scatter plots, box plots, histograms (we'll go over this tomorrow!)

## During Analysis:
1. **Use appropriate tests** - match the test to your research question and data type
2. **Consider effect sizes** - statistical significance ≠ practical significance
3. **Check residuals** - especially for regression models

## After Analysis:
1. **Interpret results substantively** - what do they mean for your research question?
2. **Consider limitations** - sample size, generalizability, causation vs. correlation
3. **Report transparently** - include all relevant statistics and confidence intervals

---

# Next Steps

Excellent work! You've now learned:

✅ **Tuesday:** R basics and data manipulation  
✅ **Wednesday:** Statistical analysis in R with communication data  

## What's Coming:
- **Thursday:** Data visualization with ggplot2

Great job building your statistical analysis skills in R! 🎉