---
title: "Tuesday - Jumpstart 2025"
author: "Dan Lane"
date: "2025-08-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(mediation)
library(summarytools)
library(modelsummary)

#load process macro from external script
source(here("PROCESS Macro","processv50","PROCESS v5.0 for R","process.R"))

#load election data
load(here("data","2020-ElectionData.rds"))
```

# Welcome to R for Communication Research!

Welcome to the second day of Jumpstart! In Group B we will be exploring mediation and moderation in R

## Today's Learning Goals:

-   Run basic moderation models
-   Visualize and probe moderation models (2 x 3-way interactions)
-   Simple slopes Analysis
-   Run mediation models (mediation and process packages)



## Get to know your dataset... data about digital media use from the 2020 U.S. Election

```{r, results='asis'}

#Dataframe summary using summarytools()
dfSummary(D1,style="multiline")


```

## The Basic Moderation Equation

```{r}



model1<-lm(SMExpPol_w1~Age_w1+Edu_w1+PolInt_w1,data=D1)

modelsummary(model1,stars=T)

```



------------------------------------------------------------------------

# Next Steps and Best Practices

Congratulations! You've now learned the fundamentals of R and data manipulation. Here are some best practices for your research:

## Project Organization Tips:

```         
your_project/
├── data/           
  ├── raw/          # Raw data files
  ├── processed/    # Clean data files
├── scripts/        # R scripts  
├── output/         # Results, plots, tables
└── README.md       # Project description
```

## Code Style Guidelines:

-   Use meaningful variable names (`trust_score` not `x1`)
-   Comment your code extensively using # 
-   Use consistent spacing and indentation
-   Break complex operations into smaller steps

## What You've Accomplished Today:

✅ **Session 1:** R basics, data types, functions\
✅ **Session 2:** Data import, tidyverse, data manipulation\
✅ **Bonus:** File organization and project management

## Coming Up:

-   Statistical analysis techniques\
-   Data visualization with ggplot2

## Resources for Continued Learning:

-   [R for Data Science](https://r4ds.hadley.nz/) - Comprehensive guide
-   [Tidyverse documentation](https://www.tidyverse.org/) - Package references
-   [RStudio Cheatsheets](https://www.rstudio.com/resources/cheatsheets/) - Quick references

**Remember:** The best way to learn R is by practicing with your own research questions and data!
