---
title: "Tuesday - Jumpstart 2025"
author: "Dan Lane"
date: "2025-08-19"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(mediation)
library(summarytools)
library(modelsummary)
library(interactions)
library(emmeans)
library(ggeffects)
library(ggplot2)
library(mediation)

#load election data
load(here("data","2020-ElectionData.rds"))
```

# Welcome to R for Communication Research!

Welcome to the second day of Jumpstart! In Group B we will be exploring mediation and moderation in R

## Today's Learning Goals:

-   Run basic moderation models
-   Visualize and probe moderation models (2 x 3-way interactions)
-   Simple slopes Analysis
-   Run mediation models (mediation and process packages)



## Get to know your dataset... data about digital media use from the 2020 U.S. Election

```{r, results='asis'}

#Dataframe summary using summarytools()
dfSummary(D1,style="multiline")



```

## The Basic Moderation Equation (2 x interaction)
### RQ1: Which combination of emotions predicts political expression social media?
### RQ2: Does age predict political expression on social media across education levels?


```{r}


fit1<-lm(SMExpPol_w1~Age_w1+Edu_w1+Anger_w1+Hope_w1,data=D1)
fit2<-lm(SMExpPol_w1~Age_w1+Edu_w1+Anger_w1*Hope_w1,data=D1)
fit3<-lm(SMExpPol_w1~Age_w1+Edu_w1+Anger_w1+Edu_w1*Age_w1,data=D1)


modlist<-list("Basic Model"=fit1,
              "Continous Moderator"=fit2,
              "Categorical Moderator"=fit3)

modelsummary(modlist,stars=T)

```

## Visualizing 2-way Interaction

```{r}


interact_plot(fit2, pred=Anger_w1,modx=Hope_w1)
interact_plot(fit3, pred=Age_w1,modx=Edu_w1)
```

## Johnson-Neyman Plots

```{r}

interactions::johnson_neyman(fit2, Anger_w1,Hope_w1)

```

## Simple Slopes

```{r}


sim_slopes(fit2, Anger_w1,Hope_w1)
sim_slopes(fit3, Age_w1,Edu_w1)
```
## Conducting Three-way Interactions

```{r}

fit4<-lm(SMExpPol_w1~Age_w1+Edu_w1+Anger_w1+Hope_w1+PolInt_w1+Conflict_1_w1,data=D1)
fit5<-lm(SMExpPol_w1~Age_w1+Edu_w1+Anger_w1*Hope_w1*Conflict_1_w1+PolInt_w1,data=D1)
summary(fit5)

modlist<-list("Basic Model"=fit4,
"3-way interaction"=fit5)

modelsummary(modlist,stars=T)

```
## Visualizing 3-way Interaction

```{r}

interact_plot(fit5, pred=Anger_w1,modx=Hope_w1,mod2=Conflict_1_w1)

```

## Extra: Interactions in Factorial Experiments

```{r}


# Set seed for reproducibility
set.seed(123)

# Simulate data for a factorial experiment with an interaction
n <- 300
treatment <- as.factor(sample(c("Control", "Treatment"), n, replace = TRUE))
condition <- as.factor(sample(c("Low", "High"), n, replace = TRUE))

# Define the model with an interaction:
# score = 50 + 10*(treatment=="Treatment") + 15*(condition=="High") + 15*(treatment=="Treatment" & condition=="High") + error
score <- 50 + 10 * (treatment == "Treatment") + 15 * (condition == "High") + 15 * (treatment == "Treatment" & condition == "High") + rnorm(n, sd = 10)

data <- data.frame(treatment, condition, score)



#Fit standard model
model <- lm(score ~ treatment * condition, data = data)
summary(model)

# The terms argument specifies the variables to use for prediction.
# The first variable (`condition`) goes on the x-axis.
# The second variable (`treatment`) is used to create different lines.
predictions <- ggpredict(model, terms = c("condition", "treatment"))
plot(predictions) # A quick plot from ggeffects

#Or In Interactions pacakge (preferred)

cat_plot(model,treatment,condition)

#Probe further... marginal means 

# Calculate the estimated marginal means for the interaction
emm_model <- emmeans(model, ~ treatment * condition)
emm_model

# Compare treatment levels within each condition
pairs(emm_model, by = "condition")

# You can also compare condition levels within each treatment
pairs(emm_model, by = "treatment")



```

## Basic mediation w/ the mediation() package

```{r}

# This creates a logical vector 'complete_rows'
complete_rows <- complete.cases(D1)

# You can then use this vector to subset the data frame.
D2 <- D1[complete_rows, ]


#Basic Regression Equation

fit6<-lm(PolInt_w1~Age_w1+Edu_w1+Anger_w1+Hope_w1,data=D2)
fit7<-lm(SMExpPol_w1~Age_w1+Edu_w1+Anger_w1+Hope_w1+PolInt_w1,data=D2)
  
modlist<-list("Mediator"=fit6,
"Predictor"=fit7)

modelsummary(modlist,stars=T)

med1<-mediate(model.m = fit6,
              model.y=fit7,
              treat="Anger_w1",
                mediator="PolInt_w1",
                sims=500,
              boot = T)

summary (med1)

```

## Moderated Mediation

```{r}


#Test for moderated mediation when age is 20 vs. 60
 test.modmed(med1, covariates.1 = list(Age_w1 = 20),covariates.2 = list(Age_w1 = 90), sims = 100)


```
## Extra: Mediation with PROCESS


```{r}

#load process macro from external script
source(here("PROCESS Macro","processv50","PROCESS v5.0 for R","process.R"))

#Regular Mediation
proc1<-process (data = D2, 
         y = "SMExpPol_w1", 
         x = "Anger_w1", 
         m ="PolInt_w1",
         cov=c("Age_w1","Hope_w1"),
         model = 4) 

#Regular Moderated
proc2<-process (data = D2, 
         y = "SMExpPol_w1", 
         x = "Anger_w1", 
         m ="PolInt_w1",
         w="Age_w1",
         cov=c("Hope_w1"),
         model = 14) 


```