---
title: "Wednesday - Jumpstart 2025"
author: "Dan Lane"
date: "2025-08-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(mediation)
library(summarytools)
library(modelsummary)
library(kableExtra)
library(apaTables)

#load election data
load(here("data","2020-ElectionData.rds"))
```

# Welcome to R for Communication Research!

Welcome to the third day of Jumpstart! In Group B we will be explore how to output r code and models

## Today's Learning Goals:

-   Basics of R Markdown
-   Refersher on summarizing data and outputting descriptives
-   Model Output using modelsummary() package
-   Output using APA Tables package

# What is R Markdown?

R Markdown is a file format that combines **Markdown text** with **R code chunks** to create dynamic documents. This very document is written in R Markdown!

# Basic Syntax

## Text Formatting

You can make text **bold** with `**bold**` or *italic* with `*italic*`.

Create lists:
- Item 1
- Item 2
  - Sub-item A
  - Sub-item B

Numbered lists:
1. First item
2. Second item
3. Third item

## Headers

Use `#` for headers. More `#` symbols = smaller headers:

```
# Header 1
## Header 2  
### Header 3
```

# Code Chunks

The real power of R Markdown is combining text with code. Here's how:

## Basic Code Chunk

This chunk shows R code and its output:

```{r basic-example}
# Simple calculation
x <- c(1, 2, 3, 4, 5)
mean(x)

# Create a simple dataset
data <- data.frame(
  category = c("A", "B", "C"),
  value = c(10, 15, 8)
)
print(data)
```

## Code Chunk Options

You can control what shows up in your output:

### Show Code Only (no output)
```{r code-only, eval=FALSE}
# This code shows but doesn't run
summary(mtcars)
plot(mtcars$mpg, mtcars$hp)
```

### Show Output Only (hide code)
```{r output-only, echo=FALSE}
# This code runs but doesn't show
head(mtcars, 3)
```

### Run Code Silently (neither shows)
```{r silent, include=FALSE}
# This runs but shows neither code nor output
secret_calculation <- 2 + 2
```

The secret calculation result was `r secret_calculation`.

## Creating Tables

### Simple Table with kable()
```{r simple-table}
kable(head(mtcars, 5), caption = "First 5 rows of mtcars dataset")
```

### Manual Table in Markdown

| Column 1 | Column 2 | Column 3 |
|----------|----------|----------|
| Value A  | 10       | TRUE     |
| Value B  | 20       | FALSE    |
| Value C  | 30       | TRUE     |

# Creating Plots

R Markdown makes it easy to include plots:

```{r basic-plot, fig.width=8, fig.height=5}
# Basic scatter plot
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  labs(title = "Miles per Gallon vs Horsepower",
       x = "Miles per Gallon",
       y = "Horsepower",
       color = "Cylinders") +
  theme_minimal()
```

## Controlling Figure Size

You can control plot dimensions with chunk options:
- `fig.width=6, fig.height=4` - sets size in inches
- `fig.align='center'` - centers the plot

```{r small-plot, fig.width=5, fig.height=3, fig.align='center'}
barplot(table(mtcars$cyl), 
        main = "Distribution of Cylinders",
        xlab = "Number of Cylinders",
        ylab = "Count",
        col = c("lightblue", "lightgreen", "lightcoral"))
```

# Inline Code

You can include R results directly in text using backticks: The mtcars dataset has `r nrow(mtcars)` observations and `r ncol(mtcars)` variables. The average MPG is `r round(mean(mtcars$mpg), 2)`.

# Useful Chunk Options

Here's a reference table of common chunk options:

```{r chunk-options-table, echo=FALSE}
options_df <- data.frame(
  Option = c("echo", "eval", "include", "message", "warning", "fig.width", "fig.height"),
  Default = c("TRUE", "TRUE", "TRUE", "TRUE", "TRUE", "7", "7"),
  Description = c(
    "Show the code",
    "Run the code",
    "Include chunk in output",
    "Show messages",
    "Show warnings", 
    "Figure width in inches",
    "Figure height in inches"
  )
)

kable(options_df, caption = "Common R Markdown Chunk Options")
```

# Mathematical Expressions

R Markdown supports LaTeX math notation:

Inline math: The formula for standard deviation is $\sigma = \sqrt{\frac{\sum(x_i - \mu)^2}{N}}$.

Display math (centered):
$$
\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}
$$

# Output Formats

You can change the output format in the YAML header:

```yaml
output:
  html_document    # Web page
  pdf_document     # PDF (requires LaTeX)
  word_document    # Microsoft Word
```

# Global Options Example

The setup chunk at the top sets global options:

```{r show-setup, eval=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,        # Show code by default
  message = FALSE,    # Hide messages
  warning = FALSE,    # Hide warnings
  fig.width = 8,      # Default figure width
  fig.height = 6      # Default figure height
)
```

# Quick Reference

## Essential Chunk Options
- `{r chunk-name}` - Basic chunk
- `{r chunk-name, echo=FALSE}` - Hide code, show output
- `{r chunk-name, eval=FALSE}` - Show code, don't run
- `{r chunk-name, include=FALSE}` - Run code, show nothing

## Knitting
- **Ctrl+Shift+K** (Windows/Linux) or **Cmd+Shift+K** (Mac) to knit
- Or click the "Knit" button in RStudio

## Pro Tips
1. Name your chunks for easier debugging
2. Use `include=FALSE` for setup chunks
3. Set global options in your setup chunk
4. Use inline code for dynamic text
5. Check your document frequently while writing

---

**Congratulations!** You now know the basics of R Markdown. This entire tutorial was written in R Markdown, demonstrating everything it teaches. Happy documenting!

## Review of how to use summarytools package

### 1. freq() - Frequency Tables for Categorical Variables

The `freq()` function creates detailed frequency tables for categorical variables.

```{r freq-basic,results="asis"}
# Frequency table for gender
freq(D1$Gender_w1, 
     main = "Gender Distribution (Wave 1)",
     style = 'rmarkdown')
```


### Customized Frequency Table

```{r freq-custom,results="asis"}
# More detailed frequency table with custom options
freq(D1$ParAct_w1, 
     main = "Political Participation - Wave 1",
     style = 'rmarkdown',
     report.nas = TRUE,      # Include missing values
     totals = TRUE,          # Show totals
     display.type = TRUE,    # Display variable type
     headings = TRUE)        # Show column headings
```

### 2. descr() - Descriptive Statistics for Numerical Variables

The `descr()` function provides comprehensive descriptive statistics for numerical variables.

### Single Variable Description

```{r descr-single,results="asis"}
# Descriptive statistics for age
descr(D1$Age_w1, 
      main = "Age Distribution - Wave 1",
      style = 'rmarkdown')
```

### Multiple Variables

```{r descr-multiple,results="asis"}
# Select news consumption variables for wave 1
news_vars_w1 <- D1 %>% 
  dplyr::select(News_TV_w1, News_Online_w1, News_Print_w1, News_Audio_w1)

descr(news_vars_w1,
      main = "News Consumption Statistics - Wave 1",
      style = 'rmarkdown')
```

### Customized Descriptive Statistics

```{r descr-custom,results="asis"}
# Political attitudes with custom statistics
political_vars <- D1 %>% 
  dplyr::select(PolInt_w1, PolEff_w1, Hope_w1, Anger_w1)

descr(political_vars,
      main = "Political Attitudes - Descriptive Statistics",
      stats = c("mean", "sd", "min", "max", "med", "IQR"),
      style = 'rmarkdown',
      transpose = TRUE,       # Transpose for better readability
      headings = TRUE)
```

### Grouped Descriptive Statistics

```{r descr-grouped,results="asis"}
# Descriptive statistics by gender
stby(D1$PolInt_w1, 
   D1$Gender_w1, 
   descr,
   main = "Political Interest by Gender",
   style = 'rmarkdown')
```

## 3. dfSummary() - Comprehensive Data Frame Summary

The `dfSummary()` function provides a complete overview of your entire dataset.

### Full Dataset Summary

```{r dfsummary-full,results="asis"}
# Complete summary of the dataset
dfSummary(D1, 
          main = "Complete Dataset Summary",
          style = 'grid',
          method = 'render',
          graph.magnif = 0.75,    # Scale down graphs
          valid.col = TRUE,       # Show valid percentages
          na.col = TRUE,          # Show missing value info
          varnumbers = FALSE,     # Don't show variable numbers
          headings = TRUE)
```

### Subset Summary

```{r dfsummary-subset,results="asis"}
# Summary of wave 1 variables only
wave1_vars <- D1 %>% 
  dplyr::select(ends_with("_w1"))

dfSummary(wave1_vars,
          main = "Wave 1 Variables Summary",
          style = 'grid',
          method = 'render',
          graph.magnif = 0.6,
          tmp.img.dir = "/tmp",   # Specify temp directory for images
          max.distinct.values = 10) # Limit distinct values shown
```

### Customized dfSummary

```{r dfsummary-custom,results="asis"}
# Focus on key demographic and political variables
key_vars <- D1 %>% 
  dplyr::select(Age_w1, Gender_w1, Edu_w1, PolInt_w1, PolEff_w1, ParAct_w1)

dfSummary(key_vars,
          main = "Key Variables Summary",
          method = 'render',
          style = 'grid',
          plain.ascii = FALSE,
          graph.col = TRUE,       # Include frequency graphs
          valid.col = TRUE,
          na.col = TRUE,
          col.widths = c(150, 100, 300, 100, 100, 100)) # Custom column widths
```

## Advanced Tips for summarytools in R Markdown Output

### 1. Print with Custom Options

```{r advanced-print,results="asis"}
# Create summary and print with specific options
my_summary <- dfSummary(D1 %>% dplyr::select(Age_w1, Gender_w1, PolInt_w1))

print(my_summary, 
      method = 'render',
      table.classes = 'table table-striped table-bordered',
      bootstrap.css = FALSE,
      escape = FALSE)
```

### 2. Save Summaries to HTML

```{r save-html, eval=FALSE}
# Save frequency table to HTML file
freq_table <- freq(D1$Gender_w1, style = 'rmarkdown')
print(freq_table, file = "gender_frequency.html")

# Save descriptive stats to HTML
desc_stats <- descr(D1$Age_w1, style = 'rmarkdown')
print(desc_stats, file = "age_descriptives.html")
```

### 3. Comparative Analysis Across Waves

```{r wave-comparison,results="asis"}
# Compare political participation across waves
par_comparison <- D1 %>% 
  dplyr::select(ParAct_w1, ParAct_w2) %>%
  dfSummary(main = "Political Participation: Wave 1 vs Wave 2",
            style = 'multiline')

print(par_comparison)
```

#### Best Practices for SummaryTools in R Markdown

1. **Set global options** in your setup chunk for consistent styling
2. **Use `results='asis'`** in chunk options for proper HTML rendering
3. **Customize table classes** for Bootstrap styling if using HTML output
4. **Control graph size** with `graph.magnif` parameter
5. **Handle missing values** explicitly with `na.col = TRUE`
6. **Use meaningful titles** with the `main` parameter
7. **Group related variables** for focused analysis

#### Styling Tips

- Use `style = 'multiline'` for all functions
- Set `bootstrap.css = FALSE` in global options
- Adjust `graph.magnif` to control plot sizes
- Use `transpose = TRUE` for better readability of descriptive statistics
- Customize column widths with `col.widths` parameter

# Outputting regression models

## Introduction to modelsummary

The `modelsummary()` function creates beautiful, publication-ready tables from statistical models. It supports dozens of model types and multiple output formats (HTML, LaTeX, Word, etc.).

## Basic Model and Data

```{r data-model}
# Assuming your data D1 is loaded
# For demonstration, we'll create sample data with similar structure

# Your regression model
fit1 <- lm(SMExpPol_w1 ~ Age_w1 + Edu_w1 + Anger_w1 + Hope_w1, data = D1)
summary(fit1)
```

# Basic modelsummary Usage

## Simple Model Table

```{r basic-table}
# Basic model summary with significance stars
modelsummary(fit1, stars = TRUE)
```

## Customizing the Table

```{r custom-table}
# More customized version
modelsummary(fit1,
             stars = TRUE,                    # Show significance stars
             title = "Predicting Social Media Political Exposure",
             notes = "Data source: Political survey Wave 1")
```

# Advanced Customization

## Custom Coefficient Names

```{r custom-names}
# Create cleaner variable names
modelsummary(fit1,
             stars = TRUE,
             coef_rename = c(
               "(Intercept)" = "Intercept",
               "Age_w1" = "Age (Wave 1)",
               "Edu_w1" = "Education Level",
               "Anger_w1" = "Political Anger",
               "Hope_w1" = "Political Hope"
             ),
             title = "Predictors of Social Media Political Exposure")
```

## Selecting Statistics

```{r custom-stats}
# Customize which statistics to show
modelsummary(fit1,
             stars = TRUE,
             statistic = c("std.error", "p.value"),  # Show SE and p-values
             gof_map = c("nobs", "r.squared", "adj.r.squared", "rmse"),
             title = "Model with Custom Statistics")
```

## Multiple Output Formats

```{r output-formats}
# Different table styles for different outputs
modelsummary(fit1,
             stars = TRUE,
             output = "kableExtra",           # Enhanced HTML tables
             title = "Enhanced HTML Table") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

# Comparing Multiple Models

```{r multiple-models}
# Create additional models for comparison
fit1 <- lm(SMExpPol_w1 ~ Age_w1 + Edu_w1 + Anger_w1 + Hope_w1, data = D1)
fit2 <- lm(SMExpPol_w1 ~ Age_w1 + Edu_w1, data = D1)
fit3 <- lm(SMExpPol_w1 ~ Anger_w1 + Hope_w1, data = D1)

# Compare models side by side
modelsummary(list("Full Model" = fit1,
                  "Demographics Only" = fit2, 
                  "Emotions Only" = fit3),
             stars = TRUE,
             title = "Model Comparison: Predictors of Social Media Political Exposure")
```

## Model Comparison with Custom Statistics

```{r model-comparison-stats}
# Compare models with AIC/BIC
modelsummary(list("Model 1" = fit1, "Model 2" = fit2, "Model 3" = fit3),
             stars = TRUE,
             gof_map = c("nobs", "r.squared", "adj.r.squared", "AIC", "BIC"),
             title = "Model Comparison with Information Criteria")
```

# Customizing Appearance

## Professional Styling

```{r professional-styling}
modelsummary(fit1,
             stars = TRUE,
             coef_rename = c(
               "(Intercept)" = "Intercept",
               "Age_w1" = "Age",
               "Edu_w1" = "Education", 
               "Anger_w1" = "Political Anger",
               "Hope_w1" = "Political Hope"
             ),
             title = "Social Media Political Exposure Model",
             notes = c("Standard errors in parentheses",
                      "* p < 0.1, ** p < 0.05, *** p < 0.01"),
             gof_map = c("nobs", "r.squared", "adj.r.squared"),
             fmt = 3,
             output = "gt")
```

## Adding Custom Rows

```{r custom-rows}
# Add custom information to the table
modelsummary(fit1,
             stars = TRUE,
             add_rows = data.frame(
               term = c("Sample", "Wave"),
               "Model 1" = c("Political Survey", "Wave 1")
             ),
             title = "Model with Additional Information")
```

# Advanced Features

## Confidence Intervals Instead of Standard Errors

```{r confidence-intervals}
modelsummary(fit1,
             statistic = "conf.int",  # Show 95% confidence intervals
             title = "Model with Confidence Intervals")
```

## Custom Significance Thresholds

```{r custom-significance}
modelsummary(fit1,
             stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01, '****' = 0.001),
             title = "Custom Significance Levels")
```

## Standardized Coefficients

```{r standardized}
library(lm.beta)

# Calculate standardized coefficients
fit1_std <- lm.beta(fit1)

modelsummary(fit1_std,
             stars = TRUE,
             title = "Standardized Coefficients Model")
```

# Exporting Tables

## Save to Different Formats

```{r export-tables, eval=FALSE}
# Save to Word document
modelsummary(fit1, 
             stars = TRUE,
             output = "table.docx",
             title = "Social Media Political Exposure Model")

# Save to LaTeX
modelsummary(fit1,
             stars = TRUE, 
             output = "table.tex",
             title = "Model for LaTeX Document")

# Save to Excel
modelsummary(fit1,
             stars = TRUE,
             output = "table.xlsx", 
             title = "Model Results")
```

# Real-World Example: Publication-Ready Table

```{r publication-ready}
# Create a publication-ready table
publication_table <- modelsummary(
  list("Social Media Political Exposure" = fit1),
  stars = c('†' = 0.1, '*' = 0.05, '**' = 0.01, '***' = 0.001),
  coef_rename = c(
    "(Intercept)" = "Constant",
    "Age_w1" = "Age",
    "Edu_w1" = "Education Level", 
    "Anger_w1" = "Political Anger",
    "Hope_w1" = "Political Hope"
  ),
  gof_rename = c(
    "nobs" = "N",
    "r.squared" = "R²",
    "adj.r.squared" = "Adj. R²",
    "rmse" = "RMSE"
  ),
  title = "Regression Results: Predictors of Social Media Political Exposure (Wave 1)",
  notes = list(
    "Standard errors in parentheses.",
    "† p < 0.1, * p < 0.05, ** p < 0.01, *** p < 0.001"
  ),
  fmt = fmt_decimal(digits = 3)
)

publication_table
```

# Summary: Key modelsummary() Arguments

| Argument | Purpose | Example |
|----------|---------|---------|
| `stars` | Add significance stars | `stars = TRUE` |
| `coef_rename` | Rename coefficients | `coef_rename = c("Age_w1" = "Age")` |
| `title` | Add table title | `title = "My Model"` |
| `notes` | Add footnotes | `notes = "Sample size = 200"` |
| `statistic` | Choose statistics to display | `statistic = "std.error"` |
| `gof_map` | Select goodness-of-fit stats | `gof_map = c("nobs", "r.squared")` |
| `output` | Set output format | `output = "kableExtra"` |
| `fmt` | Number formatting | `fmt = 3` |

# Best Practices

1. **Always use meaningful variable names** with `coef_rename`
2. **Add descriptive titles and notes** for context
3. **Choose appropriate statistics** for your audience
4. **Use consistent formatting** across tables in a document
5. **Consider your output format** (HTML, PDF, Word) when styling

The `modelsummary()` function makes it easy to create professional, publication-ready regression tables with minimal code while offering extensive customization options for advanced users.

# Introduction to apaTables

The `apaTables` package creates publication-ready tables that follow American Psychological Association (APA) formatting guidelines.


```{r descriptives-basic}
# Select key variables for descriptive table
descriptive_vars <- D1 %>% 
  dplyr::select(Age_w1, Edu_w1, PolInt_w1, PolEff_w1, SMExpPol_w1, 
         Hope_w1, Anger_w1, ParAct_w1)


```


## Basic Correlation Table

```{r correlation-basic}
# Create correlation matrix for key variables
correlation_vars <- D1 %>% 
  dplyr::select(Age_w1, Edu_w1, PolInt_w1, PolEff_w1, SMExpPol_w1, 
         Hope_w1, Anger_w1, ParAct_w1)

# APA correlation table
apa.cor.table(correlation_vars,
              filename = here("output","correlations.doc"),
              table.number = 2)
```

## Correlation Table with Custom Names

```{r correlation-custom}
# Correlation table with cleaner variable names
correlation_clean <- D1 %>% 
  dplyr::select(
    "Age" = Age_w1,
    "Education" = Edu_w1, 
    "Pol. Interest" = PolInt_w1,
    "Pol. Efficacy" = PolEff_w1,
    "SM Exposure" = SMExpPol_w1,
    "Hope" = Hope_w1,
    "Anger" = Anger_w1,
    "Participation" = ParAct_w1
  )

apa.cor.table(correlation_clean,
              here("output","correlations_clean.doc"), 
              table.number = 2,
              show.conf.interval = TRUE,  # Show confidence intervals
              landscape = TRUE)           # Landscape orientation for wide tables
```

# 3. Regression Table

## Single Regression Model

```{r regression-single}
# APA regression table for your model
apa.reg.table(fit1,
              filename=here("output","regression.doc"),
              table.number = 3)
```


# 4. Specialized Tables

## ANOVA Table

```{r anova-table}

# ANOVA model
anova_model <- aov(SMExpPol_w1 ~ Edu_w1, data = D1)

# APA ANOVA table
apa.aov.table(anova_model,
              here("output","anova.doc"),
              table.number = 4)
```



# 5. Customization Options

## Table Appearance Options

```{r customization-options}
# Comprehensive example with many options
political_vars <- D1 %>% 
  dplyr::select(
    "Political Interest" = PolInt_w1,
    "Political Efficacy" = PolEff_w1,
    "Political Hope" = Hope_w1,
    "Political Anger" = Anger_w1,
    "SM Political Exposure" = SMExpPol_w1,
    "Political Participation" = ParAct_w1
  )

apa.cor.table(political_vars,
              here("output","political_correlations.doc"),
              table.number = 6,
              show.conf.interval = TRUE,
              show.sig.stars = TRUE,
              landscape = TRUE)
```



## Different Output Formats

```{r output-formats1, eval=FALSE}
# Save to different formats
apa.cor.table(correlation_clean,
              filename = "table.doc")     # Word document (default)

apa.cor.table(correlation_clean,
              filename = "table.rtf")     # RTF format

# For LaTeX users
apa.cor.table(correlation_clean,
              filename = "table.tex")     # LaTeX format
```






